---
title: "Chapter 6"
author: "Alan T. Arnholt"
date: 'Last compiled: `r format(Sys.time(), "%B %d, %Y at %X")`'
output: bookdown::html_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning = FALSE, message = FALSE, fig.align = "center")
```

# Scatterplots, Association, and Correlation

Objectives:


I.    Examine scatterplots for association

II.   Calculate the correlation coefficient

III.  Properties of $r$

VI.   Introduce `tidyverse` (graphs, and data munging)

A **scatterplot** is a graph of points on a horizontal $x$-axis and vertical $y$-axis.  It helps us picture associations between variables.

Typically, we will have an **explanatory** or predictor variable called $x$ and a **response** variable called $y$.  These largely depend on how a problem is worded.  What you might want to predict is $y$.

To describe a scatterplot, we discuss:

1. direction (positive or negative)
2. form (straight, curved, other)
3. strength (amount of scatter - strong, moderate, weak) and
4. outliers.

A numerical measure that gives us direction and strength of a linear relationship is correlation, $r = \frac{\sum{z_xz_y}}{n-1}$.

$$r = \frac{1}{n-1}\sum_{i=1}^{n}\left(\frac{x - \bar{x}}{s_x}\right)\left(\frac{y - \bar{y}}{s_y}\right)$$

Before we calculate correlation, we must check 3 things:

1. Quantitative variables?
2. Straight (cloud shape), no curve
3. No outliers

**Properties of $r$**

1. sign of $r$ gives direction of association
2. $-1 \leq r \leq 1$
3. $r_{x,y} = r_{y,x}$
4. Correlation has no units
5. Correlation is not affected by multiplying or shifting data
6. Correlation measures LINEAR association only
7. Outliers affect correlation greatly.


# Scatterplots and the `tidyverse`

Read in the data set `Heights_and_weights.csv` stored in the `DATA` directory using the `read.csv()` function and store the results in `hw`.

```{r}
hw <- read.csv("./DATA/Heights_and_weights.csv")
head(hw) # shows first 6 rows of hw
library(tidyverse) # loads many useful packages including ggplot2 and dplyr
p <- ggplot(data = hw, aes(x = Height, y = Weight)) +
  geom_point()
#
p
p + theme_bw() 
p + theme_bw() + labs(x = "Height (inches)", y = "Weight (lbs)")
```

```{r, label = "US", fig.cap = "Scatterplot of Weight (pounds) versus Height (inches)"}
p3 <- ggplot(data = hw, aes(x = Height, y = Weight)) +
  geom_point(color = "blue") + 
  theme_bw() +
  labs(x = "Height (inches)", y = "Weight (pounds)")
p3
```

## Using `mutate()`

Use `mutate()` to create a weight variable in kg, and a height variable in cm.  

```{r}
hw <- hw %>% 
  mutate(Weight_kg = Weight*0.453592, Height_cm = Height*2.54)
head(hw)
```

Create a scatterplot similar to Figure \@ref(fig:US) using `Weight_kg` and `Height_cm` on the $y$ and $x$ axes instead of `Weight` and `Height`.

```{r, label = "EU", fig.cap = "Scatterplot of Weight (kg) versus Height (cm)"}
# Your code goes here





```

Use `mutate()` to create `zx` and `zy` as well as the variable `CC` to indicate whether the product of `zx` times `zy` is positive or negative.

```{r}
hw <- hw %>% 
  mutate(zx = (Height - mean(Height))/sd(Height), zy = (Weight - mean(Weight))/sd(Weight), 
         CC = ifelse(abs(zx*zy) <.01, "neutral", ifelse(zx*zy > 0, "positive", "negative")))
head(hw)
ggplot(data = hw, aes(x = zx, y = zy, color = CC)) +
  geom_point() +
  geom_hline(yintercept = 0, lty = "dashed") + 
  geom_vline(xintercept = 0, lty = "dashed") + 
  theme_bw()+ 
  labs(x = "Standardized Height", y = "Standardized Weight") + 
  scale_color_manual(values= c("red", "blue", "darkgreen"))
```

## Computing the correlation

Compute the correlation between `Weight` and `Height` using the information in `hw`.

```{r}
head(hw)
# Using tidyverse
hw %>% 
  summarize(r_US = cor(Height, Weight), r_EU = cor(Height_cm, Weight_kg)) 
# Using base R
cor(hw$Weight, hw$Height)
```

## Straightening Scatterplots

Read in the data in `F-stops.csv` and create a scatterplot to resemble Figure 6.11 on page 183 from the text book.

```{r}
# Your Code Goes Here






```

Recreate Figure 6.13 on page 185 of the text.

```{r}
# Your Code Goes Here




```

Compute the correlation for the variables used in Figure 6.11 and Figure 6.13 of the text book.

```{r}
# Your Code Goes Here



```

# Reading in data from homework/Pearson --- One way to do it

1. Click open in Excel and the data will download in an `*.xlsx` format  in your downloads folder.  

2. Upload the `*.xslx` file to a `HWK_DATA` directory by clicking `upload` and navigating to the file you want to upload.

```{r}
library(readxl)
prob1 <- read_xlsx("./HWK_DATA/prob1.xlsx")
knitr::kable(prob1)
```

```{r}
ggplot(data = prob1, aes(x = Number_of_Sales_People_Working,
                         y = Sales_in_Thousands)) + 
  geom_point(color = "blue") + 
  theme_bw()
```

## Problem 15 {-}

```{r}
library(readxl)
prob15 <- read_xlsx("./HWK_DATA/prob15.xlsx")
knitr::kable(prob15)
```

```{r}
ggplot(data = prob15, aes(x = `Position Number`, y = `Distance from star (million miles)`)) + 
  geom_point() + 
  theme_bw()
```

